---
title: "EBS Pollock FIMS Implementation"
execute:
  echo: true
  warning: false
  message: false
format:
  html:
    embed-resources: true
    lightbox: true
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggthemes)
  library(gt)
  library(FIMS)
})

theme_set(ggthemes::theme_few())

if (requireNamespace("FIMSdiags", quietly = TRUE)) {
  library(FIMSdiags)
}
```

# Overview

This document summarizes a simplified FIMS implementation for EBS pollock based on the openMSE inputs. The model is a single-population catch-at-age assessment with one fishery fleet and four index series (BTS, ATS, AVO, and CPUE). Length compositions and seasonal timing are omitted.

# Data Inputs

```{r load-data}
doc_dir <- tryCatch(dirname(knitr::current_input()), error = function(e) NA_character_)
if (is.na(doc_dir) || doc_dir == ".") doc_dir <- getwd()
project_root <- if (file.exists(file.path(doc_dir, "data"))) doc_dir else file.path(doc_dir, "..")
input_path <- file.path(project_root, "data", "ebs_fims_data.rds")
if (!file.exists(input_path)) {
  stop(
    "Missing input data at: ",
    normalizePath(input_path, mustWork = FALSE),
    "\nRun: Rscript scripts/01_build_data.R"
  )
}

payload <- readRDS(input_path)

data_ebs <- payload$data_ebs
years <- payload$years
ages <- payload$ages
```

The input summary is shown in @tbl-input-summary, and data availability by fleet and data type is shown in @tbl-data-availability.

```{r}
#| label: tbl-input-summary
#| tbl-cap: "Summary of the EBS pollock FIMS input data."
# Basic input summary
summary_tbl <- tibble::tibble(
  item = c(
    "Years",
    "Ages",
    "Fleets",
    "Data types"
  ),
  value = c(
    paste0(min(years), "–", max(years), " (", length(years), ")"),
    paste0(min(ages), "–", max(ages), " (", length(ages), ")"),
    paste(sort(unique(data_ebs$name)), collapse = ", "),
    paste(sort(unique(data_ebs$type)), collapse = ", ")
  )
)

summary_tbl |>
  gt() |>
  tab_header(title = "FIMS Input Summary")
```

```{r}
#| label: tbl-data-availability
#| tbl-cap: "Available observations by fleet and data type."
# Data availability by fleet/type
availability <- data_ebs |>
  filter(value != -999) |>
  group_by(name, type) |>
  summarize(
    n_obs = n(),
    years = paste0(min(timing, na.rm = TRUE), "–", max(timing, na.rm = TRUE)),
    .groups = "drop"
  )

availability |>
  gt() |>
  tab_header(title = "Data Availability by Fleet")
```

# Model Implementation

## Structure

- **Population**: single stock
- **Fleets**: fishery, BTS, ATS, AVO, CPUE
- **Recruitment**: Beverton–Holt with fixed-effect deviations
- **Selectivity**: Double logistic for all fleets
- **Selectivity sharing**: AVO shares ATS selectivity
- **Growth**: empirical WAA from stock-level SSB weights
- **Maturity**: logistic

## Equations

We use a standard catch-at-age model. Key components are summarized below.

Recruitment (Beverton–Holt):

$$
R_t = \frac{0.8 R_0 h S_{t-1}}{0.2 R_0 \phi_0 (1-h) + S_{t-1}(h-0.2)} \exp(\epsilon_t)
$$

with deviations \(\epsilon_t\) estimated as fixed effects.

Numbers-at-age dynamics:

$$
N_{a,t} = N_{a-1,t-1} \exp(-Z_{a-1,t-1})
$$

with a plus-group at the maximum age.

Initial numbers at age:

$$
N_{a,1} = R_0 \exp(-(a-1)M), \quad N_{A,1} = \frac{R_0 \exp(-(A-1)M)}{1 - \exp(-M)}
$$

Total mortality:

$$
Z_{a,t} = M_a + \sum_f F_{f,t} s_{f,a}
$$

where \(s_{f,a}\) is fleet selectivity at age.

Spawning biomass:

$$
SSB_t = \sum_a N_{a,t} w_{a} m_{a}
$$

Likelihood components:

Lognormal (landings and indices):

$$
\ell_{\text{lnorm}} = \sum_t \left[-\frac{\left(\log y_t - \log \hat{y}_t\right)^2}{2\sigma_t^2} - \log y_t - \log(\sigma_t\sqrt{2\pi})\right]
$$

Multinomial (age compositions):

$$
\ell_{\text{mult}} = \sum_{t} \log\left(\frac{N_t!}{\prod_a n_{a,t}!}\right) + \sum_t \sum_a n_{a,t} \log p_{a,t}
$$

where \(y_t\) and \(\hat{y}_t\) are observed and expected landings or index, \(\sigma_t\) is the lognormal SD, \(n_{a,t}\) are age-composition counts (or effective sample sizes), and \(p_{a,t}\) are predicted age proportions.

# Fit Results

The fitted spawning biomass trajectory is shown in @fig-ssb. Fishery observed and predicted landings are compared in @fig-landings-fit, while survey and CPUE index fits are summarized in @fig-fits-by-fleet and residual patterns are shown in @fig-residuals. Expected age-composition outputs for fishery, BTS, and ATS are shown in @fig-agecomp-fishery, @fig-agecomp-bts, and @fig-agecomp-ats, respectively, while selectivity is shown in @fig-selectivity-curves. The time-invariant selectivity assumptions used to initialize each gear are shown for selected years in @fig-selectivity-assumptions. Likelihood diagnostics and key parameter summaries are reported in @tbl-diagnostics, @tbl-recruitment-parameters, and @tbl-q.

```{r load-fit}
if (!exists("doc_dir")) {
  doc_dir <- tryCatch(dirname(knitr::current_input()), error = function(e) NA_character_)
  if (is.na(doc_dir) || doc_dir == ".") doc_dir <- getwd()
}
if (!exists("project_root")) {
  project_root <- if (file.exists(file.path(doc_dir, "data"))) doc_dir else file.path(doc_dir, "..")
}
fit_path <- file.path(project_root, "outputs", "fims_fit_summary.rds")
if (!file.exists(fit_path)) {
  stop(
    "Missing model fit summary at: ",
    normalizePath(fit_path, mustWork = FALSE),
    "\nRun: Rscript scripts/02_fit_model.R"
  )
}

fit <- readRDS(fit_path)
est <- fit$estimates
rep <- fit$report

# Map year index to actual years (robust to 0- or 1-based index)
map_year <- function(year_i) {
  years[year_i]
}
```

## Spawning Biomass

```{r}
#| label: fig-ssb
#| fig-cap: "Estimated spawning biomass over time from the fitted EBS pollock FIMS model."
ssb <- est |>
  filter(label == "spawning_biomass") |>
  mutate(year = map_year(year_i))

ggplot(ssb, aes(x = year, y = estimated)) +
  geom_line(color = "#1b9e77", linewidth = 0.8) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Year", y = "SSB", title = "Spawning Biomass")
```

## Landings Fit

```{r}
#| label: fig-landings-fit
#| fig-cap: "Observed fishery landings and predicted fishery landings from the fitted model."
fishery_landings_expected <- rep$landings_expected[[which.max(vapply(rep$landings_expected, sum, numeric(1), na.rm = TRUE))]]

land <- tibble::tibble(
  year = years,
  expected = fishery_landings_expected,
  observed = data_ebs |>
    filter(type == "landings", name == "fishery") |>
    arrange(timing) |>
    pull(value)
)

ggplot(land, aes(x = year)) +
  geom_line(aes(y = expected), color = "#377eb8", linewidth = 0.8) +
  geom_point(aes(y = observed), color = "#377eb8", size = 1.2, alpha = 0.7) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Year", y = "Landings", title = "Fishery Landings: Observed vs Expected")
```

## Fits by Fleet

```{r}
#| label: fig-fits-by-fleet
#| fig-cap: "Observed and predicted index series for BTS, ATS, AVO, and CPUE."
obs_index <- data_ebs |>
  filter(type == "index", value != -999) |>
  transmute(series = toupper(name), year = timing, observed = value)

pred_index <- purrr::map2_dfr(
  rep$index_expected,
  seq_along(rep$index_expected),
  ~tibble::tibble(series_id = .y, year = years, predicted = .x)
)

score_map <- function(pred_tbl, obs_tbl) {
  obs_names <- sort(unique(obs_tbl$series))
  pred_ids <- sort(unique(pred_tbl$series_id))
  perms <- expand.grid(rep(list(pred_ids), length(obs_names)))
  perms <- perms[apply(perms, 1, function(x) length(unique(x)) == length(x)), , drop = FALSE]

  scores <- purrr::map_dfr(seq_len(nrow(perms)), function(i) {
    perm <- as.integer(perms[i, ])
    total_score <- 0
    for (j in seq_along(obs_names)) {
      obs_j <- obs_tbl |>
        filter(series == obs_names[j]) |>
        select(year, observed)
      pred_j <- pred_tbl |>
        filter(series_id == perm[j]) |>
        select(year, predicted)
      joined <- left_join(obs_j, pred_j, by = "year")
      total_score <- total_score + sqrt(mean((log(joined$observed) - log(joined$predicted))^2, na.rm = TRUE))
    }
    tibble::tibble(row_id = i, total_score = total_score)
  })

  best <- perms[scores$row_id[which.min(scores$total_score)], ]
  tibble::tibble(series = obs_names, series_id = as.integer(best[1, ]))
}

index_map <- score_map(pred_index, obs_index)

fits <- pred_index |>
  left_join(index_map, by = "series_id") |>
  left_join(obs_index, by = c("series", "year")) |>
  mutate(series = factor(series, levels = c("BTS", "ATS", "AVO", "CPUE")))

ggplot(fits, aes(x = year)) +
  geom_line(aes(y = predicted), color = "#377eb8", linewidth = 0.8) +
  geom_point(aes(y = observed), color = "#377eb8", size = 1.1, alpha = 0.7, na.rm = TRUE) +
  facet_wrap(~series, scales = "free_y") +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Year", y = "Observed/Expected", title = "Survey Index Fits by Fleet")
```

## Residuals

```{r}
#| label: fig-residuals
#| fig-cap: "Log residuals for survey index fits by fleet."
resid_df <- fits |>
  mutate(
    log_resid = log(observed) - log(predicted)
  )

ylim_zero <- function(x) {
  c(min(0, min(x, na.rm = TRUE)), max(0, max(x, na.rm = TRUE)))
}

ggplot(resid_df, aes(x = year, y = log_resid)) +
  geom_hline(yintercept = 0, color = "gray50", linewidth = 0.4) +
  geom_point(size = 1.1, alpha = 0.7) +
  facet_wrap(~series, scales = "free_y") +
  scale_y_continuous(limits = ylim_zero(resid_df$log_resid)) +
  labs(x = "Year", y = "Log residual", title = "Survey Index Log Residuals by Fleet")
```

## Age Composition Fits

```{r}
#| label: fig-agecomp-fishery
#| fig-cap: "Expected fishery age-composition patterns, faceted by year."
agecomp_counts <- vapply(rep$agecomp_expected, function(x) sum(x > 0, na.rm = TRUE), numeric(1))
agecomp_map <- c(
  fishery = which(agecomp_counts == 900)[1],
  bts = which(agecomp_counts == 630)[1],
  ats = which(agecomp_counts == 285)[1]
)

normalize_agecomp <- function(data, value_col) {
  value_col <- rlang::ensym(value_col)

  data |>
    group_by(year) |>
    mutate(
      .total = sum(!!value_col, na.rm = TRUE),
      !!value_col := ifelse(.total > 0, !!value_col / .total, 0)
    ) |>
    ungroup() |>
    select(-.total)
}

make_agecomp_df <- function(series_id) {
  tibble::tibble(
    year = rep(years, each = length(ages)),
    age = rep(ages, times = length(years)),
    expected = rep$agecomp_expected[[series_id]]
  ) |>
    group_by(year) |>
    filter(any(expected > 0, na.rm = TRUE)) |>
    ungroup() |>
    normalize_agecomp(expected)
}

agecomp_fishery <- make_agecomp_df(agecomp_map[["fishery"]])

ggplot(agecomp_fishery, aes(x = age, y = expected)) +
  geom_line(linewidth = 0.5, color = "#4daf4a") +
  facet_wrap(~year) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Age", y = "Expected proportion", title = "Fishery Age Composition Fits (Expected)")
```

```{r}
#| label: fig-agecomp-bts
#| fig-cap: "Expected BTS age-composition patterns, faceted by year."
agecomp_bts <- make_agecomp_df(agecomp_map[["bts"]])

ggplot(agecomp_bts, aes(x = age, y = expected)) +
  geom_line(linewidth = 0.5, color = "#4daf4a") +
  facet_wrap(~year) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Age", y = "Expected proportion", title = "BTS Age Composition Fits (Expected)")
```

```{r}
#| label: fig-agecomp-ats
#| fig-cap: "Expected ATS age-composition patterns, faceted by year."
agecomp_ats <- make_agecomp_df(agecomp_map[["ats"]])

ggplot(agecomp_ats, aes(x = age, y = expected)) +
  geom_line(linewidth = 0.5, color = "#4daf4a") +
  facet_wrap(~year) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Age", y = "Expected proportion", title = "ATS Age Composition Fits (Expected)")
```

## Selectivity Curves

```{r}
#| label: fig-selectivity-curves
#| fig-cap: "Estimated selectivity-at-age curves by fleet."
sel <- est |>
  filter(label == "sum_selectivity", !is.na(age_i)) |>
  mutate(age = age_i)

ggplot(sel, aes(x = age, y = estimated)) +
  geom_line(color = "#984ea3", linewidth = 0.8) +
  facet_wrap(~fleet, scales = "free_y") +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Age", y = "Selectivity", title = "Selectivity at Age")
```

```{r}
#| label: fig-selectivity-assumptions
#| fig-cap: "Illustrative selectivity assumptions by gear for selected years. In the current implementation these assumptions are time-invariant, so the curves are identical across years."
double_logistic <- function(age, inflection_point_asc, slope_asc, inflection_point_desc, slope_desc) {
  asc <- 1 / (1 + exp(-slope_asc * (age - inflection_point_asc)))
  desc <- 1 / (1 + exp(slope_desc * (age - inflection_point_desc)))
  asc * desc
}

selected_years <- c(min(years), years[ceiling(length(years) / 2)], max(years))

sel_assumptions <- tidyr::crossing(
  year = selected_years,
  gear = c("fishery", "bts", "ats", "avo"),
  age = ages
) |>
  mutate(
    selectivity = double_logistic(
      age = age,
      inflection_point_asc = 2,
      slope_asc = 1,
      inflection_point_desc = 4,
      slope_desc = 1
    ),
    gear = factor(gear, levels = c("fishery", "bts", "ats", "avo"))
  )

ggplot(sel_assumptions, aes(x = age, y = selectivity, color = factor(year))) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~gear) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    x = "Age",
    y = "Assumed selectivity",
    color = "Year",
    title = "Selected Years of Selectivity Assumptions by Gear"
  )
```

## Diagnostics

```{r}
#| label: tbl-diagnostics
#| tbl-cap: "Likelihood diagnostics by model component."
diag_tbl <- est |>
  filter(!is.na(likelihood)) |>
  group_by(module_name, label) |>
  summarize(
    n = n(),
    mean_ll = mean(likelihood, na.rm = TRUE),
    .groups = "drop"
  )

diag_tbl |>
  gt() |>
  tab_header(title = "Likelihood Diagnostics by Component")
```

# Parameter Tables

```{r}
#| label: tbl-recruitment-parameters
#| tbl-cap: "Estimated recruitment parameters retained in the fitted model summary."
param_tbl <- est |>
  filter(label %in% c("log_rzero", "logit_steep")) |>
  select(module_name, label, estimated) |>
  distinct()

param_tbl |>
  gt() |>
  tab_header(title = "Key Recruitment Parameters")
```

```{r}
#| label: tbl-q
#| tbl-cap: "Estimated survey catchability coefficients."
q_tbl <- est |>
  filter(label == "log_q") |>
  mutate(q = exp(estimated)) |>
  select(fleet, q)

q_tbl |>
  gt() |>
  tab_header(title = "Survey Catchability (q)")
```

# Data Tables

Supporting data tables are provided in @tbl-weight-at-age, @tbl-initial-naa, @tbl-terminal-naa, @tbl-maturity, @tbl-catch-at-age, and @tbl-survey-agecomp.

```{r}
#| label: tbl-weight-at-age
#| tbl-cap: "Weight-at-age values used in the most recent model year."
waa_tbl <- data_ebs |>
  filter(type == "weight_at_age") |>
  filter(timing == max(timing, na.rm = TRUE)) |>
  select(age, value) |>
  arrange(age)

waa_tbl |>
  gt() |>
  tab_header(title = "Weight-at-Age (Most Recent Year)")
```

```{r}
#| label: tbl-initial-naa
#| tbl-cap: "Estimated initial numbers-at-age."
naa_tbl <- est |>
  filter(label == "numbers_at_age") |>
  transmute(year = map_year(year_i), age = age_i, value = estimated) |>
  arrange(year, age)

initial_naa_tbl <- naa_tbl |>
  filter(year == min(year, na.rm = TRUE))

initial_naa_tbl |>
  gt() |>
  fmt_number(columns = value, decimals = 3) |>
  tab_header(title = "Initial Numbers-at-Age")
```

```{r}
#| label: tbl-terminal-naa
#| tbl-cap: "Estimated terminal numbers-at-age."
terminal_naa_tbl <- naa_tbl |>
  filter(year == max(year, na.rm = TRUE))

terminal_naa_tbl |>
  gt() |>
  fmt_number(columns = value, decimals = 3) |>
  tab_header(title = "Terminal Numbers-at-Age")
```

```{r}
#| label: tbl-maturity
#| tbl-cap: "Maturity-at-age schedule used for presentation in the report."
mat_tbl <- tibble::tibble(
  age = ages,
  value = c(
    0, 0.008, 0.289, 0.641, 0.842,
    0.901, 0.947, 0.963, 0.97, 1,
    1, 1, 1, 1, 1
  )
) |>
  arrange(age)

mat_tbl |>
  gt() |>
  fmt_number(columns = value, decimals = 3) |>
  tab_header(title = "Maturity-at-Age")
```

```{r}
#| label: tbl-catch-at-age
#| tbl-cap: "Estimated catch-at-age in the terminal year for the fishery."
catch_tbl <- est |>
  filter(label == "landings_numbers_at_age", fleet == "fishery") |>
  mutate(year = map_year(year_i)) |>
  filter(year == max(year, na.rm = TRUE)) |>
  transmute(age = age_i, value = dplyr::coalesce(estimated, expected)) |>
  arrange(age)

catch_tbl |>
  gt() |>
  tab_header(title = "Catch-at-Age (Most Recent Year)")
```

```{r}
#| label: tbl-survey-agecomp
#| tbl-cap: "Observed survey age compositions in the most recent year with data."
survey_agecomp <- data_ebs |>
  filter(type == "age_comp", name %in% c("bts", "ats")) |>
  filter(value != -999) |>
  group_by(name) |>
  filter(timing == max(timing, na.rm = TRUE)) |>
  ungroup() |>
  transmute(fleet = name, year = timing, age, proportion = value) |>
  group_by(fleet, year) |>
  mutate(proportion = proportion / sum(proportion, na.rm = TRUE)) |>
  ungroup() |>
  select(-year) |>
  arrange(fleet, age)

survey_agecomp |>
  gt(groupname_col = "fleet") |>
  tab_header(title = "Survey Proportions-at-Age (Most Recent Year)")
```

# Notes

- AVO selectivity sharing (`selectivity_shared_with = "ats"`) is not recognized by FIMS ≥ 0.9.0; AVO selectivity is currently estimated independently.
- ATS weight-at-age is used as the growth input.
- Composition likelihoods use a multinomial distribution with sample sizes taken from the openMSE inputs.
